A memory frontier for complex synapses

### 2013-10-01 NIPS/SynapseNIPS.tex
brought NIPS paper down to page limit

This results in the following SNR
%
\begin{equation}\label{eq:SNRcalc}
  \snr(t) = \sqrt{N}\prn{2 f\pot f\dep} \eq \prn{\M\pot-\M\dep} \e^{rt\frg} \w.
\end{equation}
%
The factor of $\eq$ describes the synapses being in the steady-state distribution before the memory is encoded. The factor of $(\M\pot-\M\dep)$ comes from the encoding of the memory at $t=0$, with $\wvi$ being $\pm1$ in synapses that are potentiated/depotentiated. The factor of $\e^{rt\frg}$ describes the subsequent evolution of the probability distribution, averaged over all sequences of plasticity events, and the factor of $\w$ indicates the readout via the synaptic weight.
A detailed derivation can be found in the supplementary material.
We will frequently refer to this function as the memory curve.
It can be thought of as the excess fraction of synapses (relative to equilibrium) that maintain their ideal synaptic strength at time $t$, as dictated by the stored memory at time $t=0$.

@@

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{cascade.svg}}\label{fig:cascade}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{multistate.svg}}\label{fig:multistate}
  \item\aligntop{\includegraphics[width=0.35\linewidth]{cascms.eps}}\label{fig:cascms}
 \end{myenuma}
 \end{center}
  \caption{Models of complex synapses.
  (\ref{fig:cascade}) The cascade model of \cite{Fusi2005cascade}, showing transitions between states of high/low synaptic weight (red/blue circles) due to potentiation/depression (solid red/dashed blue arrows).
  (\ref{fig:multistate}) The serial model of \cite{Leibold2008serial}.
  (\ref{fig:cascms}) The memory curves of these two models, showing the decay of the signal-to-noise ratio (to be defined in \autoref{sec:setup}) as subsequent memories are stored.
  }\label{fig:models}
\end{figure}

@@

The critical question is: what do these constraints imply about the space of achievable memory curves in \eqref{eq:SNRcalc}?
To answer this question, especially for limits on achievable memory at finite times, it will be useful to employ the eigenmode decomposition:
%
\begin{equation}\label{eq:eigendecomp}
  \frg = \sum_a -q_a \uv^a \vv^a,
  \quad
  \vv^a \uv^b = \delta_{ab},
  \quad
  \frg \uv^a = -q_a \uv^a,
  \quad
  \vv^a \frg = -q_a \vv^a.
\end{equation}
%
Here $q_a$ are the negative of the eigenvalues of the forgetting process $\frg$, $\uv^a$ are the right (column) eigenvectors and $\vv^a$ are the left (row) eigenvectors.
This decomposition allows us to write the memory curve as a sum of exponentials,
%
\begin{equation}\label{eq:SNReigen}
\begin{aligned}
  \snr(t) &= \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a},
  &\quad \text{where}&\;&
  \CI_a &= \prn{2 f\pot f\dep} \eq (\M\pot-\M\dep) \uv^a  \vv^a \w,\\&
  & \text{and}&&
  \tau_a &= \frac{1}{q_a}.
\end{aligned}
\end{equation}
%
We can then ask the question: what are the constraints on these quantities, namely initial SNR's, $\CI_a$, and time constants, $\tau_a$, for the modes,  implied by the constraints in \eqref{eq:constr}?
We will derive some of these constraints in the next section.

@@

This two state model has the equilibrium distribution $\eq=(f\dep,f\pot)$ and its flux is given by $\F_{-+} = rf\pot f\dep$.
This is maximized when $f\pot=f\dep=\half$, leading to the upper bound:
%
\begin{equation}\label{eq:max_init}
  \snr(0) \leq \sqrt{N}.
\end{equation}
%
We note that while this model has high initial SNR, it also has very fast memory decay -- with a timescale $\tau\sim\frac{1}{r}$.
As the synapse is very plastic, the initial memory is encoded very easily, but the subsequent memories also overwrite it rapidly.
This is one example of the tradeoff between optimizing memory at early versus late times.
\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\hp\aligntop{\includegraphics[width=0.15\linewidth]{max_init_pot.svg}}\label{fig:max_init_pot}\hp
  \item\aligntop{\includegraphics[width=0.15\linewidth]{max_init_dep.svg}}\label{fig:max_init_dep}\hp
  \item\hp\aligntop{\includegraphics[width=0.05\linewidth]{binary_det.svg}}\label{fig:binary_det}
  \end{myenuma}
 \end{center}
  \caption{Synaptic models that maximize initial SNR.
  (\ref{fig:max_init_pot}) For potentiation, all transitions starting from a weak state lead to a strong state, and the probabilities for all transitions leaving a given weak state sum to 1.
  (\ref{fig:max_init_pot}) Depression is similar to potentiation, but with strong and weak interchanged.
  (\ref{fig:binary_det}) The equivalent two state model, with transition probabilities under potentiation and depression equal to one.
  }\label{fig:max_init}
\end{figure}

@@

Our goal of understanding the relationship between structure and function in the space of all possible synaptic models is complicated by the fact that this space contains many different possible network topologies, encoded in the nonzero matrix elements of $\M\potdep$.  To systematically analyze this entire space, we develop an important organizing principle using the theory of first passage times in the stochastic process of forgetting, described by $\frg$.
The mean first passage time matrix, $\fptb_{ij}$, is defined as the average time it takes to reach state $j$ for the first time, starting from state $i$.
The diagonal elements are defined to be zero.

@@

In the context of complex synapses, we can define the partial sums
%
\begin{equation}\label{eq:kemenypm}
   \eta^+_i = \sum_{j\in+} \fptb_{ij} \eq_j,
   \qquad
   \eta^-_i = \sum_{j\in-} \fptb_{ij} \eq_j.
\end{equation}
%
These can be thought of as the average time it takes to reach the strong/weak states respectively.
Using these definitions, we can then impose an order on the states by arranging them in order of decreasing $\eta^+_i$ or increasing $\eta^-_i$.
Because $\eta^+_i+\eta^-_i=\eta$ is independent of $i$, the two orderings are the same.
In this order, which depends sensitively on the structure of $\M\potdep$, states later (to the right in figures below) can be considered to be more potentiated than states earlier (to the left in figures below), despite the fact that they have the same synaptic efficacy.
In essence, in this order, a state is considered to be more potentiated if the average time it takes to reach all the strong efficacy states is shorter.
We will see that synaptic models that optimize various measures of memory have an exceedingly simple structure when, and only when, their states are arranged in this order.%
\footnote{Note that we do not need to worry about the order of the $\eta^\pm_i$ changing during the optimization: necessary conditions for a maximum only require that there is no infinitesimal perturbation that increases the area. Therefore we need only consider an infinitesimal neighborhood of the model, in which the order will not change.}

@@

With the states in the order described above, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \autoref{fig:perts}, see supplementary material for details.
For example, in \autoref{fig:perts}(\ref{fig:tri_right}), for two states $i$ on the left and $j$ on the right, with $j$ being more ``potentiated" than $i$ (i.e. $\eta^+_i > \eta^+_j$), we have proven that increasing $\M\pot_{ij}$ and decreasing $\M\dep_{ij}$ leads to an increase in area.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \autoref{fig:env}(\ref{fig:sticky}),%\autoref{fig:max_area}, 
with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some perturbation to further increase the area.

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.1\linewidth]{triangular_right.svg}}\label{fig:tri_right}\hp
  \item\aligntop{\includegraphics[width=0.1\linewidth]{triangular_left.svg}}\label{fig:tri_left}\hp
  \item\aligntop{\includegraphics[width=0.2\linewidth]{shortcut.svg}}\label{fig:shortcut}
 \end{myenuma}
 \end{center}
  \caption{Perturbations that increase the area.
  (\ref{fig:tri_right}) Perturbations that increase elements of $\M\pot$ above the diagonal and decrease the corresponding elements of $\M\dep$. It can no longer be used when $\M\dep$ is lower triangular, \ie depression must move synapses to ``more depressed'' states.
  (\ref{fig:tri_left}) Perturbations that decrease elements of $\M\pot$ below the diagonal and increase the corresponding elements of $\M\dep$. It can no longer be used when $\M\pot$ is upper triangular, \ie potentiation must move synapses to ``more potentiated'' states.
  (\ref{fig:shortcut}) Perturbation that decreases ``shortcut'' transitions and increases the bypassed ``direct'' transitions. It can no longer be used when there are only nearest-neighbor ``direct'' transitions.
  }\label{fig:perts}
\end{figure}

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
The area of a linear chain model can be expressed directly in terms of its equilibrium state distribution, $\eq$, yielding the following upper bound on the area of any model with the same $\eq$ (see supplementary material):
%
\begin{equation}\label{eq:multistatearea}
  A \leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we chose $\w_k=\sgn\brk{k - \sum_j j\eq_j}$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \autoref{fig:env}(\ref{fig:sticky}).%\autoref{fig:max_area}.
This makes it very difficult to get out of the end states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This analytical result is similar to a numerical result found in \cite{Barrett2008discrete} under a slightly different information theoretic measure of memory performance.

@@

Now we can pick some fixed time, $t_0$, and maximize the SNR at that time \wrt the parameters $\brc{\CI_a,\tau_a}$, subject to the constraints above.
This always results in a single nonzero $\CI_a$; in essence, optimizing memory at a single time requires a single exponential.  The resulting optimal memory curve, along with the achieved memory at the chosen time, depends on $t_0$ as follows:
%
\begin{equation}\label{eq:env}
\begin{aligned}
  t_0 &\leq \frac{M-1}{r} &
   \implies
  \snr(t) &= \sqrt{N}\e^{-rt/(M-1)} &
   \implies
  \snr(t_0) &= \sqrt{N}\e^{-rt_0/(M-1)},\\
  t_0 &\geq \frac{M-1}{r} &
   \implies
  \snr(t) &= \frac{\sqrt{N}(M-1)\e^{-t/t_0}}{rt_0} &
   \implies
  \snr(t_0) &= \frac{\sqrt{N}(M-1)}{\e rt_0}.
\end{aligned}
\end{equation}
%
Both the initial SNR bound and the area bound are saturated at early times.
At late times, only the area bound is saturated.
The function $\snr(t_0)$, the green curve in \autoref{fig:env}(\ref{fig:envplot}), above forms a memory curve envelope with late-time power-law decay $\sim t_0\inv$.
No synaptic model can have an SNR that is greater than this at any time.
We can use this to find an upper bound on the memory lifetime, $\tau(\epsilon)$, by finding the point at which the envelope crosses $\epsilon$:
%
\begin{equation}\label{eq:lifebound}
  \tau(\epsilon) \leq \frac{\sqrt{N}(M-1)}{\epsilon\e r},
\end{equation}
%
where we assume $N>(\epsilon\e)^2$.
Intriguingly, both the lifetime and memory envelope expand linearly with the number of internal states $M$, and increase as the square root of the number of synapses $N$.

@@

\begin{figure}
 \begin{center}
 \begin{myenuma}
 \parbox[t]{0.45\linewidth}{
  \item\aligntop{\includegraphics[width=0.95\linewidth]{env.svg}}\label{fig:envplot}
  }\hp
  \parbox[t]{0.45\linewidth}{
  \item\aligntop{\includegraphics[width=0.8\linewidth]{diffjump.svg}}\label{fig:diffjump}\\[0.1\linewidth]
  \item\aligntop{\includegraphics[width=0.8\linewidth]{multistate_sticky.svg}}\label{fig:sticky}
  }
 \end{myenuma}
 \end{center}
  \caption{The memory curve envelope for $N=100$, $M=12$.
  (\ref{fig:envplot}) An upper bound on the SNR at any time is shown in green.
  The red dashed curve shows the result of numerical optimization of synaptic models with random initialization.
  The solid red curve shows the highest SNR we have found with hand designed models.
  At early times these models are of the form shown in (\ref{fig:diffjump}) with different numbers of states, and all transition probabilities equal to $1$.
  At late times they are of the form shown in (\ref{fig:sticky}) with different values of $\varepsilon$.
  The model shown in (\ref{fig:sticky}) also saturates the area bound \eqref{eq:max_area} in the limit $\varepsilon\to0$.
  }\label{fig:env}
\end{figure}

@@

We have initiated the development of  a general theory of learning and memory with complex synapses, allowing for an exploration of the entire space of complex synaptic models, rather than analyzing individual models one at a time.
In doing so, we have obtained several new mathematical results delineating the functional limits of memory achievable by synaptic complexity, and the structural characterization of synaptic dynamical systems that achieve these limits.
In particular, operating within the ideal observer framework of \cite{Fusi2005cascade,Fusi2007multistate,Barrett2008discrete}, we have shown that for a population of $N$ synapses with $M$ internal states, (a) the initial SNR of any synaptic model cannot exceed $\sqrt{N}$, and any model that achieves this bound is equivalent to a binary synapse, (b) the area under the memory curve of any model cannot exceed that of a linear chain model with the same equilibrium distribution, (c) both the area and memory lifetime of any model cannot exceed $\CO(\sqrt{N}M)$, and the model that achieves this limit has a linear chain topology, (d) we have derived an envelope memory curve in the SNR-time plane that cannot be exceeded by the memory curve of any model, and models that approach this envelope for times greater than $\CO(\frac{M}{r})$ are linear chain models, and (e) this late-time envelope is a power-law proportional to $O({\sqrt{N} M}/{rt})$, indicating that synaptic complexity can strongly enhance the limits of achievable memory.

@@

Another important consideration is the level of noise tolerance of the synapse.
Note that the stochasticity of the models is an expression of biological noise.
However, the synapses should not be allowed to optimize all of this noise away.
It would be interesting to consider such limits on the noise levels.

Finally, it would be interesting to design physiological experiments in order to perform optimal systems identification of potential Markovian dynamical systems hiding within biological synapses.
If we measured pre and post-synaptic spike trains, and also recorded changes in post-synaptic potentials to measure changes in synaptic weight, we could use hidden Markov model techniques to find the best-fit synaptic model.
Then given our theory, we could match this measured synaptic model to optimal models to infer which timescales the synapse operates on.

>>>>

This results in the following SNR
%
\begin{equation}\label{eq:SNRcalc}
  \snr(t) = \sqrt{N}\prn{2 f\pot f\dep} \eq \prn{\M\pot-\M\dep} \e^{rt\frg} \w.
\end{equation}
%
%The factor of $\eq$ describes the synapses being in the steady-state distribution before the memory is encoded. The factor of $(\M\pot-\M\dep)$ comes from the encoding of the memory at $t=0$, with $\wvi$ being $\pm1$ in synapses that are potentiated/depotentiated. The factor of $\e^{rt\frg}$ describes the subsequent evolution of the probability distribution, averaged over all sequences of plasticity events, and the factor of $\w$ indicates the readout via the synaptic weight.
A detailed derivation of this formula can be found in the supplementary material.
We will frequently refer to this function as the memory curve.
It can be thought of as the excess fraction of synapses (relative to equilibrium) that maintain their ideal synaptic strength at time $t$, as dictated by the stored memory at time $t=0$.

@@

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{cascade.svg}}\label{fig:cascade}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{multistate.svg}}\label{fig:multistate}
  \item\aligntop{\includegraphics[width=0.3\linewidth]{cascms.eps}}\label{fig:cascms}
 \end{myenuma}
 \end{center}
  \caption{Models of complex synapses.
  (\ref{fig:cascade}) The cascade model of \cite{Fusi2005cascade}, showing transitions between states of high/low synaptic weight (red/blue circles) due to potentiation/depression (solid red/dashed blue arrows).
  (\ref{fig:multistate}) The serial model of \cite{Leibold2008serial}.
  (\ref{fig:cascms}) The memory curves of these two models, showing the decay of the signal-to-noise ratio (to be defined in \autoref{sec:setup}) as subsequent memories are stored.
  }\label{fig:models}
\end{figure}

@@

The critical question is: what do these constraints imply about the space of achievable memory curves in \eqref{eq:SNRcalc}?
To answer this question, especially for limits on achievable memory at finite times, it will be useful to employ the eigenmode decomposition:
%
\begin{equation}\label{eq:eigendecomp}
  \frg = \sum_a -q_a \uv^a \vv^a,
  \quad
  \vv^a \uv^b = \delta_{ab},
  \quad
  \frg \uv^a = -q_a \uv^a,
  \quad
  \vv^a \frg = -q_a \vv^a.
\end{equation}
%
Here $q_a$ are the negative of the eigenvalues of the forgetting process $\frg$, $\uv^a$ are the right (column) eigenvectors and $\vv^a$ are the left (row) eigenvectors.
This decomposition allows us to write the memory curve as a sum of exponentials,
%
\begin{equation}\label{eq:SNReigen}
\begin{aligned}
  \snr(t) &= \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a},
%  &\quad \text{where}&\;&
%  \CI_a &= \prn{2 f\pot f\dep} \eq (\M\pot-\M\dep) \uv^a  \vv^a \w,\\&
%  & \text{and}&&
%  \tau_a &= \frac{1}{q_a}.
\end{aligned}
\end{equation}
%
where $\;\CI_a = (2 f\pot f\dep) \eq (\M\pot-\M\dep) \uv^a  \vv^a \w\;$ and $\;\tau_a = {1}/{q_a}$.
We can then ask the question: what are the constraints on these quantities, namely eigenmode initial SNR's, $\CI_a$, and time constants, $\tau_a$, implied by the constraints in \eqref{eq:constr}?
We will derive some of these constraints in the next section.

@@

This two state model has the equilibrium distribution $\eq=(f\dep,f\pot)$ and its flux is given by $\F_{-+} = rf\pot f\dep$.
This is maximized when $f\pot=f\dep=\half$, leading to the upper bound:
%
\begin{equation}\label{eq:max_init}
  \snr(0) \leq \sqrt{N}.
\end{equation}
%
We note that while this model has high initial SNR, it also has very fast memory decay -- with a timescale $\tau\sim\frac{1}{r}$.
As the synapse is very plastic, the initial memory is encoded very easily, but the subsequent memories also overwrite it rapidly.
This is one example of the tradeoff between optimizing memory at early versus late times.
\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\hp\aligntop{\includegraphics[width=0.12\linewidth]{max_init_pot.svg}}\label{fig:max_init_pot}\hp
  \item\aligntop{\includegraphics[width=0.12\linewidth]{max_init_dep.svg}}\label{fig:max_init_dep}\hp
  \item\hp\aligntop{\includegraphics[width=0.05\linewidth]{binary_det.svg}}\label{fig:binary_det}
  \end{myenuma}
 \end{center}
  \caption{Synaptic models that maximize initial SNR.
  (\ref{fig:max_init_pot}) For potentiation, all transitions starting from a weak state lead to a strong state, and the probabilities for all transitions leaving a given weak state sum to 1.
  (\ref{fig:max_init_pot}) Depression is similar to potentiation, but with strong and weak interchanged.
  (\ref{fig:binary_det}) The equivalent two state model, with transition probabilities under potentiation and depression equal to one.
  }\label{fig:max_init}
\end{figure}

@@

Our goal of understanding the relationship between structure and function in the space of all possible synaptic models is complicated by the fact that this space contains many different possible network topologies, encoded in the nonzero matrix elements of $\M\potdep$.  To systematically analyze this entire space, we develop an important organizing principle using the theory of first passage times in the stochastic process of forgetting, described by $\frg$.
The mean first passage time matrix, $\smash{\fptb_{ij}}$, is defined as the average time it takes to reach state $j$ for the first time, starting from state $i$.
The diagonal elements are defined to be zero.

@@

In the context of complex synapses, we can define the partial sums
%
\begin{equation}\label{eq:kemenypm}
   \eta^+_i = \sum_{j\in+} \fptb_{ij} \eq_j,
   \qquad
   \eta^-_i = \sum_{j\in-} \fptb_{ij} \eq_j.
\end{equation}
%
These can be thought of as the average time it takes to reach the strong/weak states respectively.
Using these definitions, we can then impose an order on the states by arranging them in order of decreasing $\smash{\eta^+_i}$ or increasing $\smash{\eta^-_i}$.
Because $\smash{\eta^+_i+\eta^-_i=\eta}$ is independent of $i$, the two orderings are the same.
In this order, which depends sensitively on the structure of $\M\potdep$, states later (to the right in figures below) can be considered to be more potentiated than states earlier (to the left in figures below), despite the fact that they have the same synaptic efficacy.
In essence, in this order, a state is considered to be more potentiated if the average time it takes to reach all the strong efficacy states is shorter.
We will see that synaptic models that optimize various measures of memory have an exceedingly simple structure when, and only when, their states are arranged in this order.%
\footnote{Note that we do not need to worry about the order of the $\eta^\pm_i$ changing during the optimization: necessary conditions for a maximum only require that there is no infinitesimal perturbation that increases the area. Therefore we need only consider an infinitesimal neighborhood of the model, in which the order will not change.}

@@

With the states in the order described above, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \autoref{fig:perts}, see supplementary material for details.
For example, in \autoref{fig:perts}(\ref{fig:tri_right}), for two states $i$ on the left and $j$ on the right, with $j$ being more ``potentiated" than $i$ (i.e. $\smash{\eta^+_i > \eta^+_j}$), we have proven that increasing $\smash{\M\pot_{ij}}$ and decreasing $\smash{\M\dep_{ij}}$ leads to an increase in area.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \autoref{fig:env}(\ref{fig:sticky}),%\autoref{fig:max_area},
with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some perturbation to further increase the area.

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.07\linewidth]{triangular_right.svg}}\label{fig:tri_right}\hp
  \item\aligntop{\includegraphics[width=0.07\linewidth]{triangular_left.svg}}\label{fig:tri_left}\hp
  \item\aligntop{\includegraphics[width=0.14\linewidth]{shortcut.svg}}\label{fig:shortcut}
 \end{myenuma}
 \end{center}
  \caption{Perturbations that increase the area.
  (\ref{fig:tri_right}) Perturbations that increase elements of $\M\pot$ above the diagonal and decrease the corresponding elements of $\M\dep$. It can no longer be used when $\M\dep$ is lower triangular, \ie depression must move synapses to ``more depressed'' states.
  (\ref{fig:tri_left}) Perturbations that decrease elements of $\M\pot$ below the diagonal and increase the corresponding elements of $\M\dep$. It can no longer be used when $\M\pot$ is upper triangular, \ie potentiation must move synapses to ``more potentiated'' states.
  (\ref{fig:shortcut}) Perturbation that decreases ``shortcut'' transitions and increases the bypassed ``direct'' transitions. It can no longer be used when there are only nearest-neighbor ``direct'' transitions.
  }\label{fig:perts}
\end{figure}

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
The area of a linear chain model can be expressed directly in terms of its equilibrium state distribution, $\eq$, yielding the following upper bound on the area of any model with the same $\eq$ (see supplementary material):
%
\begin{equation}\label{eq:multistatearea}
  A \leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we chose $\w_k=\sgn[{k - \sum_j j\eq_j}]$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \autoref{fig:env}(\ref{fig:sticky}).%\autoref{fig:max_area}.
This makes it very difficult to get out of the end states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This analytical result is similar to a numerical result found in \cite{Barrett2008discrete} under a slightly different information theoretic measure of memory performance.

@@

Now we can pick some fixed time, $\smash{t_0}$, and maximize the SNR at that time \wrt the parameters $\smash{\brc{\CI_a,\tau_a}}$, subject to the constraints above.
This always results in a single nonzero $\smash{\CI_a}$; in essence, optimizing memory at a single time requires a single exponential.  The resulting optimal memory curve, along with the achieved memory at the chosen time, depends on $t_0$ as follows:
%
\begin{equation}\label{eq:env}
\begin{aligned}
  t_0 &\leq \frac{M-1}{r} &
   \implies
  \snr(t) &= \sqrt{N}\e^{-rt/(M-1)} &
   \implies
  \snr(t_0) &= \sqrt{N}\e^{-rt_0/(M-1)},\\
  t_0 &\geq \frac{M-1}{r} &
   \implies
  \snr(t) &= \frac{\sqrt{N}(M-1)\e^{-t/t_0}}{rt_0} &
   \implies
  \snr(t_0) &= \frac{\sqrt{N}(M-1)}{\e rt_0}.
\end{aligned}
\end{equation}
%
Both the initial SNR bound and the area bound are saturated at early times.
At late times, only the area bound is saturated.
The function $\snr(t_0)$, the green curve in \autoref{fig:env}(\ref{fig:envplot}), above forms a memory curve envelope with late-time power-law decay $\sim \smash{t_0\inv}$.
No synaptic model can have an SNR that is greater than this at any time.
We can use this to find an upper bound on the memory lifetime, $\tau(\epsilon)$, by finding the point at which the envelope crosses $\epsilon$:
%
\begin{equation}\label{eq:lifebound}
  \tau(\epsilon) \leq \frac{\sqrt{N}(M-1)}{\epsilon\e r},
\end{equation}
%
where we assume $N>(\epsilon\e)^2$.
Intriguingly, both the lifetime and memory envelope expand linearly with the number of internal states $M$, and increase as the square root of the number of synapses $N$.

@@

\begin{figure}
 \begin{center}
 \begin{myenuma}
 \parbox[t]{0.44\linewidth}{
  \item\aligntop{\includegraphics[width=0.95\linewidth]{env.svg}}\label{fig:envplot}
  }\hp
  \parbox[t]{0.45\linewidth}{
  \item\aligntop{\includegraphics[width=0.8\linewidth]{diffjump.svg}}\label{fig:diffjump}\\[0.1\linewidth]
  \item\aligntop{\includegraphics[width=0.8\linewidth]{multistate_sticky.svg}}\label{fig:sticky}
  }
 \end{myenuma}
 \end{center}
  \caption{The memory curve envelope for $N=100$, $M=12$.
  (\ref{fig:envplot}) An upper bound on the SNR at any time is shown in green.
  The red dashed curve shows the result of numerical optimization of synaptic models with random initialization.
  The solid red curve shows the highest SNR we have found with hand designed models.
  At early times these models are of the form shown in (\ref{fig:diffjump}) with different numbers of states, and all transition probabilities equal to $1$.
  At late times they are of the form shown in (\ref{fig:sticky}) with different values of $\varepsilon$.
  The model shown in (\ref{fig:sticky}) also saturates the area bound \eqref{eq:max_area} in the limit $\varepsilon\to0$.
  }\label{fig:env}
\end{figure}

@@

We have initiated the development of  a general theory of learning and memory with complex synapses, allowing for an exploration of the entire space of complex synaptic models, rather than analyzing individual models one at a time.
In doing so, we have obtained several new mathematical results delineating the functional limits of memory achievable by synaptic complexity, and the structural characterization of synaptic dynamical systems that achieve these limits.
In particular, operating within the ideal observer framework of \cite{Fusi2005cascade,Fusi2007multistate,Barrett2008discrete}, we have shown that for a population of $N$ synapses with $M$ internal states, (a) the initial SNR of any synaptic model cannot exceed $\sqrt{N}$, and any model that achieves this bound is equivalent to a binary synapse, (b) the area under the memory curve of any model cannot exceed that of a linear chain model with the same equilibrium distribution, (c) both the area and memory lifetime of any model cannot exceed $\CO(\sqrt{N}M)$, and the model that achieves this limit has a linear chain topology with only nearest neighbor transitions, (d) we have derived an envelope memory curve in the SNR-time plane that cannot be exceeded by the memory curve of any model, and models that approach this envelope for times greater than $\CO(\frac{M}{r})$ are linear chain models, and (e) this late-time envelope is a power-law proportional to $\smash{\CO({\sqrt{N} M}/{rt})}$, indicating that synaptic complexity can strongly enhance the limits of achievable memory.

@@

%Another important consideration is the tolerance of the synapses to biological noise, which could be investigated by imposing lower limits on the level of stochasticity of the models.
Finally, it would be interesting to design physiological experiments in order to perform optimal systems identification of potential Markovian dynamical systems hiding within biological synapses,
given measurements of pre and post-synaptic spike trains along with changes in post-synaptic potentials.
Then given our theory, we could match this measured synaptic model to optimal models to infer which timescales the synapse operates on.
